# Reproducible Research: Peer Assessment 1

Throughout the report I make sure that I always include the code that I use to 
generate the output by setting echo = TRUE. I also set tidy = FALSE 
to be able to create nicer line breaks.
```{r set_global_options, echo = TRUE}
opts_chunk$set(echo = TRUE, tidy = FALSE)
```

[Making inline code visible]
(http://stackoverflow.com/questions/20409172/how-to-display-verbatim-inline-r-code-with-backticks-using-rmarkdown):
```{r make_inline_code_visible}
rinline <- function(code) {
  sprintf('[code: ``` `r %s` ```] %s', deparse(substitute(code)), code)
}
```
Demonstration of inline code:
```{r define_something}
a <- 3
```
Show the inline code and its result: `r rinline(a)`.

## Obtaining the repository including the data

1. Go to the [repository for this course]
(https://github.com/rdpeng/RepData_PeerAssessment1).

2. Click fork [when I was doing this (2014-06-11), the latest commit SHA-1 was dc20c7c4e92aea6af318a611e2669d5b9c6cbfaf].

3. On the local machine execute the command 
git clone https://github.com/rdpeng/RepData_PeerAssessment1.

4. Navigate to the directory RepData_PeerAssessment1 
using setwd("your_directory"):
```{r}
setwd(".")
```

## Loading and preprocessing the data

### Load the data (i.e. read.csv())

Set the zip filename:
```{r set_zip_filename}
filename <- 'activity.zip'
```
Unzip activity.zip:
```{r unzip_zip_file}
unzip(filename)
```
Set the data-set filename:
```{r set_csv_filename}
filename <- 'activity.csv'
```
Read the file into a data frame called activity:
```{r get_activity_data}
column_classes <- c("integer","Date","integer")
activity <- read.csv(filename,colClasses = column_classes)
```

### Process/transform the data (if necessary) into a format suitable for your analysis



Inspect the data:
```{r inspect_data}
head(activity)
str(activity)
```
We can see the variables:

- steps: Number of steps taken in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

We can also see that we got all 17,568 observations.

There should be
`r rinline(24*60/5)`
intervals in a day. Look at the relevant section of the data
```{r end_of_day}
activity[285:291,]
```
The intervals are given in the form 830 meaning 8:30 a.m. and not 830 minutes.
To plot these times with the correct scale,
I convert these to minutes since the start of the day:
```{r convert_intervals_to_minutes}
activity[,"interval"] <- (floor(activity[,"interval"]/100)*60 
                          + activity[,"interval"] %% 100)
```
Look at the end of the first day again:
```{r end_of_day_again}
activity[285:291,]
```
There are `r rinline(24*60)` minutes per day.

## What is mean total number of steps taken per day?

Get the total number of steps per day, ignoring NAs:
```{r get_total_number_of_steps_per_day}
steps_per_day <- aggregate(activity[,"steps"],by=as.list(activity["date"]),
                           FUN=sum,na.rm=TRUE)
names(steps_per_day) <- c("date","total_steps")
```
Look at the first few lines:
```{r check_total_number_of_steps_per_day}
head(steps_per_day)
```

### Make a histogram of the total number of steps taken each day

Create a histogram of the total number of steps taken per day:
```{r total_number_of_steps_per_day_histogram, fig.height = 4 }
par( mar = c(5,4,2,1) )
hist(steps_per_day$total_steps, breaks=20, 
     xlab="total number of steps per day in thousands",
     main="Histogram of total number of steps taken each day",
     xlim = c(0,22500), axes = FALSE, ylim = c(0,10) )
axis(1, at = seq(0,22000,2000), labels= seq(0,22,2))
axis(2, at = seq(0,10,2), labels = seq(0,10,2) )
```

### Calculate and report the mean and median total number of steps taken per day

Get the mean total number of steps per day
```{r mean_total_number_of_steps}
mean_total_number_of_steps <- round(mean(steps_per_day[,"total_steps"]),0)
```
The person had `r rinline(mean_total_number_of_steps)` as the mean total number 
of steps per day.  

Get the median total number of steps per day
```{r median_total_number_of_steps}
median_total_number_of_steps <- median(steps_per_day[,"total_steps"])
```
The person had `r rinline(median_total_number_of_steps)` as the median total 
number of steps per day. 

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Get the number of steps for the 5 minute intervals averaged over all days
```{r mean_steps_per_interval}
mean_steps_per_interval <- aggregate(activity[,"steps"],
                                     by = as.list(activity["interval"]),
                                     FUN = mean, na.rm=TRUE)
names(mean_steps_per_interval) <- c("interval","mean_steps")
```
Look at the first few results:
```{r check_mean_steps_per_interval}
head(mean_steps_per_interval)
```
Plot the result
```{r mean_steps_per_interval_plot, fig.height = 4}
par( mar = c(5,4,2,1))
plot(mean_steps_per_interval$interval,mean_steps_per_interval$mean_steps,
     type="l", xlab = "start time of interval", 
     ylab = "mean steps per five minute interval",
     main = "mean steps per five minute interval averaged over all days",
     axes = FALSE, xlim = c(0,1440), ylim = c(0,220))
axis(1, at = seq(0,1440,length.out=13), labels = paste0(seq(0,24,length.out=13),":00"))
axis(2, at = seq(0,225,25), labels = seq(0,225,25) )
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Get the start of the maximum interval.
```{r get_maximum_interval}
interval_start <- mean_steps_per_interval[which.max(mean_steps_per_interval[,"mean_steps"]),"interval"]
interval_start
maximum_mean_steps <- round(max(mean_steps_per_interval[,"mean_steps"]),0)
```
`r rinline(maximum_mean_steps)`, the maximum mean number of steps, occurs in 
the `r rinline(interval_start)` to 
`r rinline(interval_start+5)` minute interval. This corresponds to the interval starting at
`r rinline(floor(interval_start/60))`:`r rinline(interval_start %% 60)`.

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

Obtain the number of missing values
```{r get_number_of_missing_values}
number_of_na <- sum(is.na(activity$steps))
number_of_na
total_records <- length(activity$steps)
total_records
```
There are `r rinline(number_of_na)` NAs in the data set. 
This corresponds to `r rinline(round(number_of_na/total_records*100,1))`% missing values.

### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The imputation strategy is to replace the NAs with the mean for the interval
as calculated above. This introduces non-integer steps.
Create a function for that purpose
```{r get_mean_for_interval}
get_mean_for_interval <- function(interval){
    mean_steps_per_interval[ mean_steps_per_interval[,"interval"] == interval,
                             "mean_steps"]
}
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

Replace NA values with the mean value for that interval as calculated above:
```{r create_imputed_data_set, tidy = FALSE}
imputed_activity <- activity
imputed_activity[is.na(imputed_activity[,"steps"]), "steps"] <- 
    sapply(imputed_activity[is.na(imputed_activity[,"steps"]), "interval"],
           get_mean_for_interval)
```
Have a look at the result:
```{r check_result_of_imputation}
head(imputed_activity)
```

### Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Get the total number of steps per day
```{r get_total_number_of_steps_after_imputation}
imputed_steps_per_day <- aggregate(imputed_activity[,"steps"],
                                   by=as.list(imputed_activity["date"]),
                                   FUN=sum)
names(imputed_steps_per_day) <- c("date","total_steps")
```
```{r check_imputed_steps_per_day}
head(imputed_steps_per_day)
```
Create a histogram of the total number of steps taken per day after missing 
values were imputed:
```{r imputed_total_number_of_steps_per_day_histogram, fig.height = 4}
par(mar = c(5,4,2,2))
hist(imputed_steps_per_day$total_steps, breaks=20,
     xlab="total number of steps per day in thousands",
     main="Histogram of total number of steps taken each day",
     xlim = c(0,22500), axes = FALSE, ylim = c(0,20) )
axis(1, at = seq(0,22000,2000), labels= seq(0,22,2))
axis(2, at = seq(0,20,5), labels = seq(0,20,5) )
```

Get the mean total number of steps per day
```{r get_imputed_mean_total_number_of_steps}
imputed_mean_total_number_of_steps <- round(mean(imputed_steps_per_day[,"total_steps"]),0)
```
The person had `r rinline(sprintf("%i",imputed_mean_total_number_of_steps))` as the mean imputed total number of steps 
(compared to `r rinline(sprintf("%i",mean_total_number_of_steps))` unimputed).

Get the median total number of steps per day
```{r get_imputed_median_total_number_of_steps}
imputed_median_total_number_of_steps <- median(imputed_steps_per_day[,"total_steps"])
```
The person had `r rinline(sprintf("%i", round(imputed_median_total_number_of_steps, 0)))` 
as the median imputed total number of steps 
(compared to `r rinline(sprintf("%i", round(median_total_number_of_steps, 0)))` unimputed).

Imputation increases the estimates of the average total number of steps per day. 
The difference between the median and mean vanishes.
```{r difference_imputed_unimputed}
min(imputed_steps_per_day["total_steps"] - steps_per_day["total_steps"],na.rm=FALSE)
max(imputed_steps_per_day["total_steps"] - steps_per_day["total_steps"],na.rm=FALSE)
```
Thus the impact of the imputation strategy is to either increase or not change the 
number of steps per day.

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

Add a column to imputed_activity indicating if a day was a workday or a weekend.
Notice that weekend days (Saturday, Sunday) start with S.
```{r add_day_category}
imputed_activity["day_category"] <- "weekday"
imputed_activity[grepl("^[Ss]",weekdays(imputed_activity[,"date"])),"day_category"] <- "weekend"
```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

Get the number of steps for the 5 minute intervals averaged over all 
weekdays and weekend days
```{r get_imputed_mean_steps_per_interval}
imputed_mean_steps_per_interval <- aggregate(imputed_activity[,"steps"],
                                             by=as.list(c(imputed_activity["day_category"],
                                                          imputed_activity["interval"])),
                                             FUN=mean)
names(imputed_mean_steps_per_interval) <- c("day_category","interval","mean_steps")
```
Look at the first few observations:
```{r check_imputed_mean_steps_per_interval}
head(imputed_mean_steps_per_interval)
```
Plot the result
```{r imputed_mean_steps_per_interval_plot}
library(ggplot2)
g <- ggplot(imputed_mean_steps_per_interval,aes(interval,mean_steps))
g <- g + geom_line() + facet_wrap(~ day_category, nrow = 2) + theme_bw()
g <- g + ggtitle("Mean steps per interval split into weekdays and weekends")
g <- g + xlab("start time of five minute interval") 
g <- g + ylab("mean steps per interval")
g <- g + scale_x_continuous(breaks = seq(0, 1440, length.out = 13),
                            labels = paste0(seq(0, 24, length.out = 13),":00"))
g
```
Based on the figure, we can see that the start to the day on a weekend is a bit 
slower than on a weekday.
Further there is an additional peak between 8 and 9 p.m. on weekends. The 
imputation strategy did not differentiate between weekdays and weekends and 
might have diluted the results.
